Phylogenetics and the human microbiome.

### 2014-04-22 sbreview.tex
Tweaks before printing out.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on counts of DNA substrings of a given length, which are often called $k$-mer classifiers \citep{wang2007naive,rosen2008metagenome}.
For those studies and others, the conventional metrics of classification performance such as precision and recall are applied to data simulated from known and taxonomically classified genomes.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the phylogenetic tree.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \cite{Koslicki2013-mc}, and \cite{Droge2014-lj}.
\citet{treangen2013metamos} report shorter run times and much higher accuracy (for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.

@@

It is not straightforward to define a clear notion of optimality for OTU inference.
In phylogenetics we would like to compare reconstructions to an object that is generally not knowable---the ``true'' historical phylogenetic tree---OTU inference has a variety of desirable outcomes, only some of which are knowable.
\cite{Wang2013-ol} divide performance measures into \textit{external} measures, which compare an inferred clustering to some defined outcome, and \textit{internal} measures, which give overall descriptive statistics on inferred clustering.
External measures applied to observed or simulated data include obtaining the same number of OTUs as taxonomic groups at some level \citep[e.g.][]{edgar2013uparse}, scoring deviation from a decomposition of a phylogenetic tree \citep[e.g.][]{navlakha2009finding}, or a given set of taxonomic classifications \citep[e.g.][]{Cai2011-bi}.
Internal measures evaluate the clustering in various ways without reference to a ``true'' clustering, with the general idea that within-cluster distances should be small compared to between-cluster distances.

There are many OTU inference methods with various speeds and strategies.
Some methods proceed through a list of sequences and progressively add each either to an existing cluster or start a new cluster with that sequence, such as CD-HIT \citep{li2006cdhit} and USEARCH \citep{edgar2010usearch}, while UPARSE \citep{edgar2013uparse} uses a similar strategy while also attempting to correct for sequencing error.
\citet{white2010alignment} show that different ways of doing this genre of heuristic clustering can result in very different results.
\citet{Cai2011-bi} perform highly efficient clustering using a pseudometric based partition tree, which can be thought of as a hierarchical clustering tree with a fixed set of internal node heights.
\citet{navlakha2009finding} take a semi-supervised approach in that input includes sequences along with a subset of sequences that are equipped with taxonomic classifications; the algorithm then groups all sequences (many in novel clusters) into clusters that have similar properties as the example taxonomic clusters.
\citet{Wang2013-ol} optimize a criterion of cluster modularity.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \cite{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.

The centrality of the OTU concept can be seen by the fact that the by-sample table of OTU observations (i.e.\ the matrix of counts with rows representing samples and columns representing OTUs) is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as such errors are especially problematic there.
\citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a large number of sequencing reads and so can be used as an OTU clustering method.
With the exception of the work of \citet{sharpton2011phylotu,Zhang2013-wh}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
For example, the most commonly used index is the Simpson index \cite{simpson1949measurement}, which is simply the sum of the squared frequencies across OTUs.
The drawback of applying such a diversity metric to a collection of OTUs is that the overall structure of the diversity is lost, in that two closely related OTUs contribute as much to the diversity measure as two distant ones.
% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used and here we will focus on their applications and methods.

Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
Whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
The earth-movers distance between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'' required to move piles of dirt (each dirt pile configuration sized in proportion to one probability distribution) in one configuration to another along the tree, where work is defined as mass times distance.
In this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
There are clear connections between UniFrac and phylogenetic diversity explored by \cite{faith2009cladistic}, who point out related measures in Faith's earlier work.

@@

The large data sets associated with human microbiome analysis require highly efficient algorithms for \emph{de novo} tree inference.
Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has incorporated the clearcut code \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut, FastTree, and RAxML \cite{Stamatakis2006-yz}.
% AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
% Bayesian phylogenetic inference is absent.

@@

There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a so-called fixed \textit{reference tree}, sometimes with uncertainty estimates.
These programs \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} have various speeds and features.
This work has also spurred development of specialized alignment tools for this mapping process.
\citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
\citet{Mirarab2012-pk} use data set partitioning to improve alignments on subsets of taxa specifically for this application.
\citet{Brown2013-ia} use locality-sensitive hashing to obtain placement more than two orders of magnitude faster than the \textit{pplacer} program of \cite{matsen2010pplacer}.

@@

Developing community assembly models forms an important project for microbial ecology generally and human microbial ecology in particular, phylogeny-aware methods could be further developed.
One way to model community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
\citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \citep{mooers1997inferring,aldous2011five}.
In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics by reducing a distribution on the tips of a tree down to a real number.
Another challenge is to bring together macroevolutionary modeling with species abundance modeling, but some initial steps have been made in another setting \citep{lambert2013predicting}.

@@

In a similar vein, assessing the significance level of an observed difference between communities is not straightforward and poses difficult problems.
The randomization of group membership commonly used in combination with UniFrac to determine significance does not have appropriate properties when in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.
Such non-independence can lead to incorrect rejection of the null hypothesis.
Imagine, for example, that we have a random observation process on the tree equipped with some collection $B$ of observations at the leaves.
Each sample from the process takes a random subset from $B$ and then throws down some number of reads for each observation in that set, the number of which has mean significantly greater than 1.
If the size of $B$ is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
In trying to remedy such false-positive identification of differences, it becomes clear that even basic definitions pose a challenge: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.

For both alpha and beta diversity studies, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
One type of normalization handles differential depths of sequencing across samples.
The presently used approach is \textit{rarefaction}, which means uniform sub-sampling to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another, which is not the case.
An alternative is provided by \citet{o2012phylogenetic}, who provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly.
This is a good start, but more work should be done exploring results under deviation from that model.

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiota?
At least for the next several years we can expect the research questions described above to continue to unfold.
Future research projects will continue to bring deeper sequencing on more samples.
The uBiome \footnote{\url{http://ubiome.com/}} and American Gut \footnote{\url{http://americangut.org/}} projects promise to bring gut microbiome sequencing to the average citizen for a low price.
Comparative studies will continue to investigate what shapes and is shaped by the microbiota.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

@@

Human microbiota research has experienced a frenetic rate of expansion over the past decade, and sometimes the hype has outmatched the science.
However, our microbes are here to stay and so is research on them.
Thus we can look forward to the field of human microbiota analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
Phylogenetics has already contributed significantly to research on the human microbiota and will continue to do so.

>>>>

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on counts of DNA substrings of a given length, which are often called $k$-mer classifiers \citep{wang2007naive,rosen2008metagenome}.
For those studies and others, the conventional metrics of classification performance such as precision and recall are applied to data simulated from known and taxonomically classified genomes.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the phylogenetic tree.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \citet{Koslicki2013-mc}, and \citet{Droge2014-lj}.
\citet{treangen2013metamos} report shorter run times and much higher accuracy (for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.

@@

It is not straightforward to define a clear notion of optimality for OTU inference.
In phylogenetics we would like to compare reconstructions to an object that is generally not knowable---the ``true'' historical phylogenetic tree---OTU inference has a variety of desirable outcomes, only some of which are knowable.
\citet{Wang2013-ol} divide performance measures into \textit{external} measures, which compare an inferred clustering to some defined outcome, and \textit{internal} measures, which give overall descriptive statistics on inferred clustering.
External measures applied to observed or simulated data include obtaining the same number of OTUs as taxonomic groups at some level \citep[e.g.][]{edgar2013uparse}, scoring deviation from a decomposition of a phylogenetic tree \citep[e.g.][]{navlakha2009finding}, or a given set of taxonomic classifications \citep[e.g.][]{Cai2011-bi}.
Internal measures evaluate the clustering in various ways without reference to a ``true'' clustering, with the general idea that within-cluster distances should be small compared to between-cluster distances.

There are many OTU inference methods with various speeds and strategies.
Some methods proceed through a list of sequences and progressively add each either to an existing cluster or start a new cluster with that sequence, such as CD-HIT \citep{li2006cdhit} and USEARCH \citep{edgar2010usearch}, while UPARSE \citep{edgar2013uparse} uses a similar strategy while also attempting to correct for sequencing error.
\citet{white2010alignment} show that different ways of doing this genre of heuristic clustering can result in very different results.
\citet{Cai2011-bi} perform highly efficient clustering using a pseudometric based partition tree, which can be thought of as a hierarchical clustering tree with a fixed set of internal node heights.
\citet{navlakha2009finding} take a semi-supervised approach in that input includes sequences along with a subset of sequences that are equipped with taxonomic classifications; the algorithm then groups all sequences (many in novel clusters) into clusters that have similar properties as the example taxonomic clusters.
\citet{Wang2013-ol} optimize a criterion of cluster modularity.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.

The centrality of the OTU concept can be seen by the fact that the by-sample table of OTU observations (i.e.\ the matrix of counts with rows representing samples and columns representing OTUs) is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as such errors are especially problematic there.
\citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a large number of sequencing reads and so can be used as an OTU clustering method.
With the exception of the work of \citet{sharpton2011phylotu} and \citet{Zhang2013-wh}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
For example, the most commonly used index is the Simpson index \citep{simpson1949measurement}, which is simply the sum of the squared frequencies across OTUs.
The drawback of applying such a diversity metric to a collection of OTUs is that the large-scale structure of the diversity is lost, such that two closely related OTUs contribute as much to the diversity measure as two distant ones.
% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used and here we describe their application.

Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
Whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
The earth-movers distance between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'' required to move piles of dirt (each dirt pile configuration sized in proportion to one probability distribution) in one configuration to another along the tree, where work is defined as mass times distance.
In this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
There are clear connections between UniFrac and phylogenetic diversity explored by \citet{faith2009cladistic}, who point out related measures in Faith's earlier work.

@@

The large data sets associated with human microbiome analysis require highly efficient algorithms for \emph{de novo} tree inference.
Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has incorporated the clearcut code \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut, FastTree, and RAxML \citep{Stamatakis2006-yz}.
% AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
% Bayesian phylogenetic inference is absent.

@@

There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a so-called fixed \textit{reference tree}, sometimes with uncertainty estimates.
These programs \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} have various speeds and features.
This work has also spurred development of specialized alignment tools for this mapping process.
\citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
\citet{Mirarab2012-pk} use data set partitioning to improve alignments on subsets of taxa specifically for this application.
\citet{Brown2013-ia} use locality-sensitive hashing to obtain placement more than two orders of magnitude faster than the \textit{pplacer} program of \citet{matsen2010pplacer}.

@@

Developing community assembly models forms an important project for microbial ecology generally and human microbial ecology in particular, phylogeny-aware methods could be further developed.
One way to model community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
\citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \citep{mooers1997inferring,aldous2011five}.
In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics by reducing a distribution on the tips of a tree down to a real number.
Another challenge is to bring together macroevolutionary modeling with species abundance modeling, where some initial steps have been made in another setting \citep{lambert2013predicting}.

@@

In a similar vein, assessing the significance level of an observed difference between communities is not straightforward and poses difficult problems.
The randomization of group membership commonly used in combination with UniFrac to determine significance does not have appropriate properties when in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.
Such non-independence can lead to incorrect rejection of the null hypothesis.
Imagine, for example, that we have a random observation process on the tree equipped with some collection $B$ of observations at the leaves.
Each sample from the process takes a random subset from $B$ and then throws down some number of reads for each observation in that subset, with the number of reads having a mean significantly greater than 1.
If the size of $B$ is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
In trying to remedy such false-positive identification of differences, it becomes clear that even basic definitions pose a challenge: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.

For both alpha and beta diversity measurement, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
One type of normalization handles differential depths of sequencing across samples.
The presently used approach is \textit{rarefaction}, which means uniform sub-sampling to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another, which is not the case.
An alternative is provided by \citet{o2012phylogenetic}, who provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly.
This is a good start, but more work should be done exploring results under deviation from that model.

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiota?
At least for the next several years we can expect the research questions described above to continue to unfold.
Future research projects will continue to bring deeper sequencing on more samples.
The uBiome\footnote{\url{http://ubiome.com/}} and American Gut\footnote{\url{http://americangut.org/}} projects promise to bring gut microbiome sequencing to the average citizen for a low price.
Comparative studies will continue to investigate what shapes and is shaped by the microbiota.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

@@

Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes the hype has outmatched the science.
However, our microbes are here to stay and so is research on them.
Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
Phylogenetics has already contributed significantly to research on the human microbiome and will continue to do so.
